Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-05-01 sources/bmc_template_submission/bmc_article.tex
more tiny edits

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data provides
the opportunity to simultaneously study
the evolutionary history of many taxa in a reproducible and continuous manner.
Increased taxonomic sampling is known to improve
phylogenetic reconstructions \cite{hillis1996inferring, natsidis2019phylogenomics},
time of divergence estimates \cite{schulte2013undersampling, soares2015influence},
and biogeographic analyses \cite{kayaalp2017back},
as well as help in resolving phylogenetic conflict \cite{hedtke2006resolution, townsend2010optimal, natsidis2019phylogenomics}.
Here, we introduce Physcraper, a Python package that provides a data interoperability
framework to automate data connections across biological databases,
with the main goal of building upon published alignments and phylogenies
to extend existing phylogenetic inferences with sequence data available for more taxa
in public DNA databases.
% Physcraper updates the starting single locus alignment and corresponding phylogeny
% with public DNA data from GenBank \cite{benson2000genbank}, and links tip taxon
% labels in the updated trees to a unified, interoperable
% taxonomic resource \cite{rees2017automated}, streamlining connections between phylogenies and biological data.
Physcraper links tip taxon labels in the updated trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}, effectively streamlining connections
between phylogenies and any type of biological data.

Information such as
geographical location, fossil ranges, and genetic and phenotypic data increasingly
available in public biological databases constitute an amazing
resource for scientific discovery \cite{baxevanis2015importance}.
One of the main challenges for automatic
integration of data across biological databases are varying taxonomic idiosyncrasies.
To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for automatic taxonomic name standardization, by integrating taxonomic
data from several resources \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) taxonomy \cite{federhen_ncbi_2012, schoch_ncbi_2020},
and the Global Biodiversity Information Facility (GBIF) \cite{gbif_secretariat_gbif_2019}, among many others.
OpenTree's unified taxonomy, along with tools and methods to manage it, are available
as Application Programming Interfaces (APIs),
which are implemented as open access services for the general public \cite{opentreeAPIs}.
Physcraper leverages on existing OpenTree's unified taxonomy APIs to automatically
standardize taxon names in any phylogeny, providing
a framework for connection of updated phylogenies with data from any biological database.

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data provides
the opportunity to simultaneously study
the evolutionary history of many taxa in a reproducible and continuous manner.
Increased taxonomic sampling is known to improve
phylogenetic reconstructions \cite{hillis1996inferring, natsidis2019phylogenomics},
time of divergence estimates \cite{schulte2013undersampling, soares2015influence},
and biogeographic analyses \cite{kayaalp2017back},
as well as help in resolving phylogenetic conflict \cite{hedtke2006resolution, townsend2010optimal, natsidis2019phylogenomics}.
Here, we introduce Physcraper, a Python package that provides a data interoperability
framework to automate data connections across biological databases,
with the main goal of building upon published alignments and phylogenies
to extend existing phylogenetic inferences with sequence data available for more taxa
in public DNA databases.
% Physcraper updates the starting single locus alignment and corresponding phylogeny
% with public DNA data from GenBank \cite{benson2000genbank}, and links tip taxon
% labels in the updated trees to a unified, interoperable
% taxonomic resource \cite{rees2017automated}, streamlining connections between phylogenies and biological data.
Physcraper links tip taxon labels in the updated trees to a unified
taxonomic resource \cite{rees2017automated}, effectively streamlining connections
between phylogenies and any type of biological data.

Information such as
geographical location, fossil ranges, and genetic and phenotypic data increasingly
available in public biological databases constitute an amazing
resource for scientific discovery \cite{baxevanis2015importance}.
One of the main challenges for automatic
integration of data across biological databases are varying taxonomic idiosyncrasies.
To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for automatic taxonomic name standardization, by integrating taxonomic
data from several resources \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) taxonomy \cite{federhen_ncbi_2012, schoch_ncbi_2020},
and the Global Biodiversity Information Facility (GBIF) \cite{gbif_secretariat_gbif_2019}, among many others.
OpenTree's unified taxonomy, along with tools and methods to manage it, are available
as Application Programming Interfaces (APIs),
which are implemented as open access services for the general public \cite{opentreeAPIs}.
Physcraper leverages on existing OpenTree's unified taxonomy APIs to automatically
standardize taxon names in any phylogeny, a key step to streamline the connection
of updated phylogenies with data from different and independent biological database.
